---
title: 'cm006: `dplyr` Exercise'
output: 
 html_document:
   keep_md: true 
editor_options: 
  chunk_output_type: inline
---
<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

**Optional, but recommended startup**:

1. Change the file output to both html and md _documents_ (not notebook).

2. `knit` the document. 

3. Stage and commit the rmd, and knitted documents.

# Intro to `dplyr` syntax

Load the `gapminder` and `tidyverse` packages. Hint: `suppressPackageStartupMessages()`!
    - This loads `dplyr`, too.
    
```{r load packages, warning = FALSE, message = FALSE}
# load your packages here:
library(tidyverse)
library(gapminder)
```
    

## `select()` (8 min)

1. Make a data frame containing the columns `year`, `lifeExp`, `country` from the gapminder data, in that order.

```{r}
select(gapminder, year, lifeExp, country)
```


2. Select all variables, from `country` to `lifeExp`.

```{r}
# This will work:
select(gapminder, country, continent, year, lifeExp)

# Better way:
select(gapminder, country:lifeExp)
```


3. Select all variables, except `lifeExp`.

```{r}
select(gapminder, -lifeExp)
```

4. Put `continent` first. Hint: use the `everything()` function.

```{r}
select(gapminder, continent, everything()) 
```


5. Rename `continent` to `cont`. There are two ways to rename columns.

```{r}
# compare
select(gapminder, cont = continent, everything()) #when remaning things with select, it will rename but then select that new column in the order that you write it

rename(gapminder, cont = continent) #to rename in the same place, use the rename function
```


## `arrange()` (8 min)

1. Order by year.

```{r}
arrange(gapminder, year) #all rows with 1952 come first
```

2. Order by year, in descending order.

```{r}
arrange(gapminder, desc(year)) #now starts at the highest number & goes down to 1952 
```

3. Order by year, then by life expectancy.

```{r}
arrange(gapminder, year, lifeExp)
```

## Piping, `%>%` (8 min)

Note: think of `%>%` as the word "then"!

Demonstration:

Here I want to combine `select()` Task 1 with `arrange()` Task 3.

This is how I could do it by *nesting* the two function calls:

```{r nesting functions example, eval = F}
# Nesting function calls can be hard to read
arrange(select(gapminder, year, lifeExp, country), year, lifeExp) #wanted to select something, and then arrange something, but if you wanted to add more tasks, this line of code would become too hard to read eventually
# Piping allows you to write it out as you want it to happen 
```

Now using with pipes:

```{r}
# alter the below to include 2 "pipes"
gapminder %>%
  select(year, lifeExp, country) %>%
  arrange(year, lifeExp)
# this has a property described as self-documenting, tells you exactly what is happening without the need to write many comments 

# if you did arrange(year, country) instead it would arrange the countries alphabetically, so arrange doesn't change the order of the columns but within them 
```

# Resume lecture 

Return to guide at section 6.7.

## `filter()` (10 min)

1. Only take data with population greater than 100 million.

```{r}
gapminder %>%
  filter(pop > 100000000)
```

2. Your turn: of those rows filtered from step 1., only take data from Asia.

```{r} 
#The three have the same results

gapminder %>%
  filter(pop > 100000000) %>%
  filter(continent == "Asia") #1 

gapminder %>%
  filter(pop > 100000000 & continent == "Asia") #2

gapminder %>%
  filter(pop > 100000000,
         continent == "Asia") #3
```

3. Repeat 2, but take data from countries Brazil, and China. 

```{r}
gapminder %>%
  filter(pop > 100000000,
         country == "Brazil" | country == "China")
         
```

## `mutate()` (10 min)

Let's get: 

- GDP by multiplying GPD per capita with population, and
- GDP in billions, named (`gdpBill`), rounded to two decimals.

```{r}
gapminder %>%
  mutate(gdpBill = gdpPercap*pop/1000000000)

#there are functions like round() that allow you to round up to decimals, or significance()
0.86768
round(0.86768, digits = 2)
```

Notice the backwards compatibility! No need for loops!

Try the same thing, but with `transmute` (drops all other variables). 

```{r}
gapminder %>%
  transmute(gdpBill = gdpPercap*pop/1000000000) #using mutate & select (can be a shortcut if you want to do those two and don't want to run two lines of code)
```

The `if_else` function is useful for changing certain elements in a data frame.

Example: Suppose Canada's 1952 life expectancy was mistakenly entered as 68.8 in the data frame, but is actually 70. Fix it using `if_else` and `mutate`. 

```{r}
gapminder %>%
  mutate(lifeExp = if_else(country == "Canada" & year == "1952", 70, lifeExp)) #so this means that when the country is Canada and the year is 1952, we want the value to be 70, but we want the rest to remain the same so we just write lifeExp
```

Your turn: Make a new column called `cc` that pastes the country name followed by the continent, separated by a comma. (Hint: use the `paste` function with the `sep=", "` argument).

```{r}
gapminder %>%
  mutate(cc = paste("country", "continent"))
```

These functions we've seen are called __vectorized functions__.

## git stuff (Optional)

Knit, commit, push!

# Bonus Exercises

If there's time remaining, we'll practice with these three exercises. I'll give you 1 minute for each, then we'll go over the answer.

1. Take all countries in Europe that have a GDP per capita greater than 10000, and select all variables except `gdpPercap`. (Hint: use `-`).

2. Take the first three columns, and extract the names.

3. Of the `iris` data frame, take all columns that start with the word "Petal". 
    - Hint: take a look at the "Select helpers" documentation by running the following code: `?tidyselect::select_helpers`.
    
4. Convert the population to a number in billions.

5. Filter the rows of the iris dataset for Sepal.Length >= 4.6 and Petal.Width >= 0.5.

Exercises 3. and 5. are from [r-exercises](https://www.r-exercises.com/2017/10/19/dplyr-basic-functions-exercises/).